### 2017-11-15 Draft2/Appending.Dynamic.tex
Lots of work

It is well-known that appending one item to a dynamic array takes amortized $O(1)$ time. Thus, appending $n$ items takes $O(n)$ time. However, this fact is not very useful for the purpose of comparison--- later, it will be shown that growth arrays also take $O(n)$ time to append $n$ items.

It would be helpful if the ratio $\dfrac{\FuncTimeDynamicArray(n)}{\FuncTimeGrowthArray(n)}$ could be determined, where $\FuncTimeDynamicArray(n)$ and $\FuncTimeGrowthArray(n)$ denote the average time needed to append $n$ items for dynamic and growth arrays, respectively. This represents how much faster growth arrays are than dynamic ones. However, there is no way to quantify these functions without empirical measurements. Thus, I will approximate this ratio through a different technique: Suppose $n$ elements are appended to an empty collection. How many times is an element stored in an array? I will term the answer to this question the \textbf{write cost} of $n$ appends, and denote it $\FuncWrites(n)$.

I will na\"{i}vely assume that $\FuncTime(n) \propto \FuncWrites(n)$, or that the average time for $n$ appends is proportional to their write cost. Then $\dfrac{\FuncTimeDynamicArray(n)}{\FuncTimeGrowthArray(n)} = \dfrac{\FuncWritesDynamicArray(n)}{\FuncWritesGrowthArray(n)}$, where $\FuncWritesDynamicArray$ and $\FuncWritesGrowthArray$ are respectively the write cost functions for dynamic and growth arrays. In this section, I will derive the formula for $\FuncWritesDynamicArray(n)$. Because this section concerns dynamic arrays only, I will write it as $\FuncWrites(n)$.

In the code for $\FuncAppend$, one array store is performed unconditionally, so it is apparent that $\FuncWrites(n) \geq n$ after $n$ appends. However, $\FuncGrow$ also does some writing, so in order to find a precise formula for $\FuncWrites(n)$, I need to analyze when $\FuncGrow$ is called. To do this, I use the following lemma:

@@

\begin{proof}
	It was shown in Lemma \ref{lem:CapacitySeq} that if $n > \VarInitCapacity$, $\VarUseful = \ExprMessy \geq 1$.  Now, note that $\VarUseful$ may be written as $\ExprMessyAlt$ for such $n$. Then,
	\begin{align*}
	\ExprMessyAltInner \leq \VarUseful &< \ExprMessyAltInner + 1\\
	\frac{n}{\VarInitCapacity} \leq \VarUseful &< \frac{\VarGrowthFactor{n}}{\VarInitCapacity}
	\end{align*}
	as desired.
\end{proof}

I can build on this inequality to receive asymptotic bounds for $\FuncWrites(n)$:
\begin{align*}
\frac n \VarInitCapacity &\FluteLeq \VarGrowthFactor^\VarUseful \FluteLess \frac {\VarGrowthFactor{n}} \VarInitCapacity\\
\frac n \VarInitCapacity &\FluteLeq \VarGrowthFactor^\VarUseful - 1 \FluteLess \frac {\VarGrowthFactor{n}} \VarInitCapacity &&\text{(Corollary \ref{coro:PlusConstant}, Theorem \ref{thm:InterchangeableInInequality})}\\
\frac n {\VarGrowthFactor - 1} &\FluteLeq \left( \frac{\VarGrowthFactor^\VarUseful - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity \FluteLess \frac {\VarGrowthFactor{n}} {\VarGrowthFactor - 1} &&\text{(Theorem \ref{thm:BothSidesInequality})}\\
\left( \frac{\VarGrowthFactor}{\VarGrowthFactor - 1} \right) n &\FluteLeq \FuncWrites(n) \FluteLess \left( \frac{2\VarGrowthFactor - 1}{\VarGrowthFactor - 1} \right) n &&\text{(Theorem \ref{thm:BothSidesInequalityAdd})}
\end{align*}
\HdrSpaceComplex

@@

First, from the definition of $\VarList.\FieldCapacity$, note that a dynamic array's capacity is the length of the buffer it stores its items in. Then a buffer of length $c$ is allocated at some point if and only if $c \in \VarCapacitySeq$. Then the total length of those buffers is
\begin{align*}
\FuncSpace(n) &= \sum_i {\VarCapacitySeq_i}\\
&= \VarInitCapacity + \VarGrowthFactor\VarInitCapacity + \VarGrowthFactor^2\VarInitCapacity + \ldots + \VarGrowthFactor^\VarUseful\VarInitCapacity\\
&= \left( \frac{\VarGrowthFactor^{\VarUseful + 1} - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity
\end{align*}
Using Lemma \ref{lem:ToVarUsefulPowerBounds} again, I asymptotically bound $\FuncSpace(n)$:
\begin{align*}
\frac n \VarInitCapacity &\FluteLeq \VarGrowthFactor^\VarUseful \FluteLess \frac {\VarGrowthFactor{n}} \VarInitCapacity\\
\frac {\VarGrowthFactor n} \VarInitCapacity &\FluteLeq \VarGrowthFactor^{\VarUseful + 1} \FluteLess \frac {\VarGrowthFactor^2 n} \VarInitCapacity &&\text{(Theorem \ref{thm:BothSidesInequality})}\\
\frac {\VarGrowthFactor n} \VarInitCapacity &\FluteLeq \VarGrowthFactor^{\VarUseful + 1} - 1 \FluteLess \frac {\VarGrowthFactor^2 n} \VarInitCapacity &&\text{(Corollary \ref{coro:PlusConstant}, Theorem \ref{thm:InterchangeableInInequality})}\\
\left( \frac{\VarGrowthFactor}{\VarGrowthFactor - 1} \right) n &\FluteLeq \FuncSpace(n) \FluteLess \left( \frac{\VarGrowthFactor^2}{\VarGrowthFactor - 1} \right) n &&\text{(Theorem \ref{thm:BothSidesInequality})}
\end{align*}

>>>>

It is well-known that appending one item to a dynamic array takes amortized $O(1)$ time. Thus, appending $n$ items takes $O(n)$ time. It will be shown later, however, that growth arrays also take $O(n)$ time for this task, so this does not let us compare time complexities.

I wish to find the ratio $\dfrac{\FuncTimeDynamicArray(n)}{\FuncTimeGrowthArray(n)}$, where $\FuncTimeDynamicArray(n)$ and $\FuncTimeGrowthArray(n)$ denote the average time needed to append $n$ items for dynamic and growth arrays, respectively. This represents how much faster growth arrays are at appending than dynamic ones. I will approximate this ratio using the following question: Suppose $n$ elements are appended to an empty collection. How many times is an element stored in an array? I will term the answer to this question the \textbf{write cost} of $n$ appends, and denote it $\FuncWrites(n)$.

I will na\"{i}vely assume that $\FuncTime(n) \propto \FuncWrites(n)$, or that the average time for $n$ appends is proportional to their write cost. Then $\dfrac{\FuncTimeDynamicArray(n)}{\FuncTimeGrowthArray(n)} = \dfrac{\FuncWritesDynamicArray(n)}{\FuncWritesGrowthArray(n)}$, where $\FuncWritesDynamicArray$ and $\FuncWritesGrowthArray$ are respectively the write cost functions for dynamic and growth arrays. In this section, I will derive the formula for $\FuncWritesDynamicArray(n)$. Because this section concerns dynamic arrays only, I will denote it $\FuncWrites(n)$.

In the code for $\FuncAppend$, one array store is performed unconditionally, so it is apparent that $\FuncWrites(n) \geq n$ after $n$ appends. However, $\FuncGrow$ also does some writing via $\FuncArrayCopy$, so in order to find a precise formula for $\FuncWrites(n)$, I need to analyze when $\FuncGrow$ is called. To do this, I use the following lemma:

@@

\begin{proof}
	It was shown in Lemma \ref{lem:CapacitySeq} that if $n > \VarInitCapacity$, $\VarUseful = \ExprMessy \geq 1$.  Now, note that $\VarUseful$ may be written as $\ExprMessyAlt$ for such $n$. Then,
	\begin{align*}
	\ExprMessyAltInner &\leq \VarUseful < \ExprMessyAltInner + 1\\
	\frac{n}{\VarInitCapacity} &\leq \VarGrowthFactor^\VarUseful < \frac{\VarGrowthFactor{n}}{\VarInitCapacity}
	\end{align*}
	as desired.
\end{proof}

I can build on this inequality to receive asymptotic bounds for $\FuncWrites(n)$:
\begin{align*}
\frac n \VarInitCapacity &\FluteLeq \VarGrowthFactor^\VarUseful \FluteLess \frac {\VarGrowthFactor{n}} \VarInitCapacity\\
\frac n \VarInitCapacity &\FluteLeq \VarGrowthFactor^\VarUseful - 1 \FluteLess \frac {\VarGrowthFactor{n}} \VarInitCapacity &&\text{(Corollary \ref{coro:PlusConstant}, Theorem \ref{thm:InterchangeableInInequality})}\\
\frac n {\VarGrowthFactor - 1} &\FluteLeq \left( \frac{\VarGrowthFactor^\VarUseful - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity \FluteLess \frac {\VarGrowthFactor{n}} {\VarGrowthFactor - 1} &&\text{(Theorem \ref{thm:BothSidesInequality})}\\
\left( \frac{\VarGrowthFactor}{\VarGrowthFactor - 1} \right) n &\FluteLeq \FuncWrites(n) \FluteLess \left( \frac{2\VarGrowthFactor - 1}{\VarGrowthFactor - 1} \right) n &&\text{(Theorem \ref{thm:BothSidesInequalityAdd})}
\end{align*}
This means that, for large $n$, the number of items written to append $n$ items is roughly between the two bounds shown. For example, suppose $\VarInitCapacity = 4$, $\VarGrowthFactor = 2$, and $32$ items are appended. $\FuncGrow$ is called at the sizes $4$, $8$, and $16$, so $4 + 8 + 16 = 28$ items are copied. Counting the $32$ writes made by $\FuncAppend$, one per item, the write cost is $\FuncWrites(32) = 28 + 32 = 60$. This is approximately equal to the lower bound, $\dfrac{2}{2 - 1} \cdot 32 = 64$.

Now, suppose that one more item is added so that $n = 33$. $\FuncGrow$ is called again, so the number of items copied becomes $4 + 8 + 16 + 32 = 60$. The write cost becomes $\FuncWrites(33) = 33 + 60 = 93$. This is approximately equal to the upper bound, $\dfrac{2 \cdot 2 - 1}{2 - 1} \cdot 33 = 99$.

Intuitively, it makes sense that $\FuncWrites(n)$ should jump from the lower bound to the upper bound when $n = \VarGrowSeq_i + 1$, since at that point $O(n)$ many items are copied from the old to the new buffer. (In the previous example, $\VarGrowSeq_i + 1$ is $17$.) As new items are added, however, both $n$ and $\FuncWrites(n)$ only increase by $1$, and the ratio $\dfrac {\FuncWrites(n)} n$ slowly dwindles toward the lower bound until $n$ reaches $\VarGrowSeq_{i + 1} + 1$.

\HdrSpaceComplex

@@

First, from the definition of $\VarList.\FieldCapacity$, note that a dynamic array's capacity is the length of the buffer it stores its items in. Then a buffer of length $c$ is allocated at some point if and only if $c \in \VarCapacitySeq$. Then the total length of those buffers is
\begin{align*}
\FuncSpace(n) &= \sum_i {\VarCapacitySeq_i}\\
&= \VarInitCapacity + \VarGrowthFactor\VarInitCapacity + \VarGrowthFactor^2\VarInitCapacity + \ldots + \VarGrowthFactor^\VarUseful\VarInitCapacity\\
&= \left( \frac{\VarGrowthFactor^{\VarUseful + 1} - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity
\end{align*}
Using Lemma \ref{lem:ToVarUsefulPowerBounds} again, I asymptotically bound $\FuncSpace(n)$:
\begin{align*}
\frac n \VarInitCapacity &\FluteLeq \VarGrowthFactor^\VarUseful \FluteLess \frac {\VarGrowthFactor{n}} \VarInitCapacity\\
\frac {\VarGrowthFactor n} \VarInitCapacity &\FluteLeq \VarGrowthFactor^{\VarUseful + 1} \FluteLess \frac {\VarGrowthFactor^2 n} \VarInitCapacity &&\text{(Theorem \ref{thm:BothSidesInequality})}\\
\frac {\VarGrowthFactor n} \VarInitCapacity &\FluteLeq \VarGrowthFactor^{\VarUseful + 1} - 1 \FluteLess \frac {\VarGrowthFactor^2 n} \VarInitCapacity &&\text{(Corollary \ref{coro:PlusConstant}, Theorem \ref{thm:InterchangeableInInequality})}\\
\left( \frac{\VarGrowthFactor}{\VarGrowthFactor - 1} \right) n &\FluteLeq \FuncSpace(n) \FluteLess \left( \frac{\VarGrowthFactor^2}{\VarGrowthFactor - 1} \right) n &&\text{(Theorem \ref{thm:BothSidesInequality})}
\end{align*}
Using the same intuition from the time complexity section, $\FuncSpace(n)$ approaches the lower bound as $n \to \VarGrowSeq_i$, and jumps to the upper bound when $n = \VarGrowSeq_i + 1$. If $100$ items were appended to a dynamic array with $\VarInitCapacity = 11$ and $\VarGrowthFactor = 3$, for example, the wasted space $s(100) - 100$ would be approximately $\dfrac{3^2}{3 - 1} \cdot 100 - 100 = 350$. If $n$ were $99$, however, it would be approximately $\dfrac{3}{3 - 1} \cdot 99 - 99 = 49.5$.

### 2017-11-15 Draft2/Appending.Growth.tex
Add pictures, [WIP] title all the algos
Lots of work

In this section, $\VarList$ is a growth array. The following algorithm implements appending for growth arrays.

\begin{algorithm}[H]
	\begin{algorithmic}[1]
		\Procedure{$\FuncAppend$}{$\VarList,\ \ParamItem$}
			\If{$\VarList.\FieldFull$}
				\State $\VarList.\FuncGrow()$
			\EndIf

@@

Now, I focus on finding a formula for the write cost function, $\FuncWritesGrowthArray(n)$, which will be written as $\FuncWrites(n)$ in this section. I claim that Lemma \ref{lem:CapacitySeq} still holds, since growth arrays satisfy all properties used by that proof. In particular, although growth arrays use a different growth algorithm than dynamic arrays, the following statement still holds:

@@

Let $\FuncWritesByGrow(n)$ denote the total number of writes made by $\FuncGrow$, and let $\VarNumPointersTail$ be the size of the tail. Since Corollary \ref{coro:GrowthSeq} also holds true for growth arrays, $\FuncGrow$ is called $|\VarGrowSeq|$ times. A buffer is appended to the tail each time $\FuncGrow$ is called. Thus, the tail's size is
\begin{align*}
\VarNumPointersTail &= |\VarGrowSeq| = |\VarCapacitySeq| - 1 = \VarUseful
\end{align*}
Then the formula for $\FuncWritesByGrow(n)$ is simply $\FuncWritesTail(\VarUseful)$, since $\VarUseful$ pointers are appended to the tail, which is a dynamic array. Finally, adding the writes made by $\FuncAppend$, the formula for $\FuncWrites(n)$ is
\begin{align*}
\FuncWrites(n) = n + \FuncWritesTail(\VarUseful)
\end{align*}
Now, I approximate $\FuncWrites(n)$ using $\sim$. To do this, I will derive the big-O complexity of $\VarUseful$.

@@

Now when $\FuncWrites(n)$ is approximated with $\sim$, $\FuncWritesTail(\VarUseful)$ disappears:
\begin{align*}
\FuncWrites(n) &= n + \FuncWritesTail(\VarUseful) = n + O(\VarUseful) = n + O(\log n) \sim n
\end{align*}
The last step is justified by Theorem \ref{thm:RemovesLowerOrderTerms}.

>>>>

The following diagrams should provide intuition on how growth arrays work.

\begin{figure}[H]
	\centering{\includegraphics[width=5in]{GrowthArrayDiagram}}
	\caption{Appending $30$ items to a growth array with $\VarInitCapacity = 4$, $\VarGrowthFactor = 2$}
\end{figure}

\begin{figure}[H]
	\centering{\includegraphics[width=5in]{GrowthArrayDiagram2}}
	\caption{Appending $16$ items to a growth array with $\VarInitCapacity = 2$, $\VarGrowthFactor = 3$}
\end{figure}

\begin{figure}[H]
	\centering{\includegraphics[width=5in]{GrowthArrayDiagram3}}
	\caption{Appending $1$ item to a growth array with $\VarInitCapacity = 4, \VarGrowthFactor = 2$. Initial size: $8$}
\end{figure}

In this section, $\VarList$ denotes a growth array. The following algorithm implements appending for growth arrays.

\begin{algorithm}[H]
	\caption{Appending \TextGrowthArray}
	\begin{algorithmic}[1]
		\Procedure{$\FuncAppend$}{$\VarList,\ \ParamItem$}
			\If{$\VarList.\FieldFull$}
				\State $\VarList.\FuncGrow()$
			\EndIf

@@

Now, I focus on finding a formula for growth arrays' write cost function, $\FuncWritesGrowthArray(n)$, which will be written as $\FuncWrites(n)$ in this section. I claim that Lemma \ref{lem:CapacitySeq} also holds for growth arrays, since they satisfy all properties used by that proof. In particular, although growth arrays use a different growth algorithm than dynamic arrays, they still have the following property:

@@

Let $\FuncWritesByGrow(n)$ denote the total number of writes made by $\FuncGrow$, and let $\VarNumPointersTail$ be the size of the tail. Since Corollary \ref{coro:GrowthSeq} also holds for growth arrays, $\FuncGrow$ is called $|\VarGrowSeq|$ times. A buffer is appended to the tail each time $\FuncGrow$ is called. Thus, the tail's size is
\begin{align*}
\VarNumPointersTail &= |\VarGrowSeq| = |\VarCapacitySeq| - 1 = \VarUseful
\end{align*}
Then the formula for $\FuncWritesByGrow(n)$ is simply $\FuncWritesTail(\VarUseful)$, since $\VarUseful$ pointers are appended to the tail, which is a dynamic array. Finally, adding the writes made by $\FuncAppend$, the formula for $\FuncWrites(n)$ is
\begin{align*}
\FuncWrites(n) = n + \FuncWritesTail(\VarUseful)
\end{align*}
Now, I approximate $\FuncWrites(n)$ using $\sim$. To do this, I will derive the big-O complexity of $\VarUseful$.

@@

Now when $\FuncWrites(n)$ is approximated with $\sim$, $\FuncWritesTail(\VarUseful)$ disappears:
\begin{align*}
\FuncWrites(n) &= n + \FuncWritesTail(\VarUseful) = n + O(\VarUseful) = n + O(\log n) \sim n
\end{align*}
(The last step is justified by Theorem \ref{thm:RemovesLowerOrderTerms}.)

This says that when $n$ items are added, the number of items written is approximately $n$. This is a significant improvement over the asymptotic bounds for $\FuncWritesDynamicArray(n)$.

### 2017-11-15 Draft2/ExternalDefinitions.tex
Add pictures, [WIP] title all the algos
Lots of work

\begin{algorithm}[H]
	\begin{algorithmic}
		\LineComment{Copies $len$ items from $source$ to $dest$}
		\State $\FuncArrayCopy(source,\ dest,\ len)$
		\State
		\LineComment{Returns a new array with length $len$}
		\State $\FuncNewArray(len)$
		\State
		\LineComment{Returns a new, empty dynamic array}
		\State $\FuncNewDynamicArray()$
	\end{algorithmic}
\end{algorithm}

I will also use the following constants in my algorithms. Specific values for these constants must be supplied by the person using the algorithms.

>>>>

\begin{algorithm}[H]
	\caption{Runtime-defined functions}
	\begin{algorithmic}
		\LineComment{Copies $len$ items from $source$ to $dest$}
		\State $\FuncArrayCopy(source,\ dest,\ len)$
		\State
		\LineComment{Returns a new array with length $len$}
		\State $\FuncNewArray(len)$
		\State
		\LineComment{Returns a new, empty dynamic array}
		\State $\FuncNewDynamicArray()$
	\end{algorithmic}
\end{algorithm}

\subsection{Constants}

I will use the following constants in my algorithms. These constants govern how the algorithms will behave. Specific values for them must be supplied by the person using the algorithms.

### 2017-11-15 Draft2/Intro.tex
add -> append
Lots of work

In imperative languages, the dynamic array is the most common data structure used by programs. People often want to add multiple items to a collection and then iterate it, which dynamic arrays make simple and efficient. But can it be said they have the \textit{most} efficient algorithm for this pattern?

@@

When dynamic arrays run out of space, they allocate a new buffer, copy the old one's contents into it, then throw the old one away. Growth arrays are less wasteful in this scenario. Instead of discarding the filled buffer, they simply archive it. The new buffer they allocate represents a continuation of the items from the old buffer. For example, if the old buffer contained items $0 -- 9$, the new buffer would contain items $10$ and beyond. Because of this technique, growth arrays allocate less memory to store the same number of items, and do not need to copy the contents of old buffers.

>>>>

In imperative languages, the dynamic array is the most common data structure used by programs. People often want to append multiple items to a collection and then iterate it, which dynamic arrays make simple and efficient. But can it be said they have the \textit{most} efficient algorithm for this pattern?

@@

When dynamic arrays run out of space, they allocate a new buffer, copy the old one's contents into it, then throw the old one away. Growth arrays are less wasteful in this scenario. Instead of discarding the filled buffer, they simply archive it. The new buffer they allocate represents a continuation of the items from the old buffer. For example, if the old buffer contained items $0 - 9$, the new buffer would contain items $10$ and beyond. Because of this technique, growth arrays allocate less memory to store the same number of items, and do not need to copy the contents of old buffers.
